Designing Types for R, Empirically

### 2020-11-03 paper/main.tex
publisher edits

\begin{document}
\title{Designing Types for R, Empirically}
\author{Alexi Turcotte}\affiliation{\institution{Northeastern University}}
\author{Aviral Goel}\affiliation{\institution{Northeastern University}}
\author{Filip Krikava}\affiliation{\institution{Czech Technical University in Prague}}
\author{Jan Vitek}\affiliation{\institution{Northeastern / Czech Technical University}}

@@

\begin{abstract}
The R programming language is widely used in a variety of domains.  It
was designed to favor an interactive style of programming with minimal
syntactic and conceptual overhead. This design is well suited to data
analysis, but a bad fit for tools such as compilers or program
analyzers. In particular, R has no type annotations, and all
operations are dynamically checked at run-time. The starting point for
our work are the twin questions: \emph{what expressive power is needed
to accurately type R code?} and \emph{which type system is the R
community willing to adopt?} Both questions are difficult to answer
without actually experimenting with a type system. The goal of this
paper is to provide data that can feed into that design process. To
this end, we perform a large corpus analysis to gain insights in the
degree of polymorphism exhibited by idiomatic R code and explore
potential benefits that the R community could accrue from a simple
type system.  As a starting point, we infer type signatures for
\NUMFUNCTIONS functions from \NUMPACKAGES packages among the most
widely used open source R libraries. We then conduct an evaluation on
\NUMPKGSEVAL clients of these packages, as well as on end-user code
from the Kaggle competition.
\end{abstract} \maketitle

@@

R and its ancestor S were designed, implemented, and maintained by
statisticians. Originally they aimed to be glue languages for 
statistical routines written in Fortran. Over three decades they
became widely used across many fields for exploratory data analysis.
Modern R, as an object of study, is fascinating. It is a vectorized,
dynamically typed, lazy functional language with limited side-effects,
extensive reflective facilities and retrofitted object-oriented programming
support.

@@

The goal of this paper is to gather data that can be used as input to
the process of designing a type system for R.  To focus our work, we
chose to limit the scope of investigation on the design of a language
for function type signatures.  Thus, this paper will use the collected
data to document the signature of user-defined functions.  For this,
we design a simple type language, one that matches the R data types.
We then extract call and return types from execution traces of a
corpus of widely used libraries, and finally synthesize type
signatures.  This allows us to see how far one can get with a simple
type language and identify limitations of our design. We validate the
robustness of the extracted type signatures by implementing a contract
checker that weaves types around their respective functions, and use
clients of the target packages for validation.  To sum up, we make the
following contributions:
\begin{itemize}
\item We implemented scalable and robust tooling to automatically
  extract type signatures and instrument R functions with checks
  based on their declared types.
\item We carried out a corpus analysis of \PACKAGES widely used and
  maintained packages to synthesize function type signatures and
  validated the robustness of the signatures against \NUMFILESRUN
  programs that use those functions.
\item We report on the appropriateness and usefulness of a simple type
  language for R.
\item Our data and code are open source and publicly
  available.\footnote{ \url{github.com/PRL-PRG/propagatr},
  \url{github.com/PRL-PRG/contractr}, data: \url{doi.org/10.5281/zenodo.4091818}, artifact: \url{doi.org/10.5281/zenodo.4037278}}
\end{itemize}

@@

\subsection{Synthesizing signatures}

@@

\subsection{Types from traces}

@@

\paragraph{Missing arguments}
Parameters which receive no values when the function is called are
termed missing (not to be confused with \NA). This occurs when a
function is called with too few arguments and no default values are
specified for those missing arguments.  We record a \k{missing} type
for such arguments.  There are two obvious ways to deal with missing
arguments: type them as \any or type them as some unit type.  We
conservatively type them as \any.

\paragraph{Non-local returns}
When a function exits with a longjump, there is no return value to speak
of. To ensure call traces are valid when a longjump occurs, we intercept the
unwinding process and record a special \k{jumped} return type for function
returns that are skipped. As we cannot be sure of the intended return value, 
these \k{jumped} values become \any types.

@@

\paragraph{Implementation details.}
We primarily rely on eight callbacks: \k{closure\_entry},
\k{closure\_exit}, \k{builtin\_entry}, \k{builtin\_exit},
\k{special\_entry}, \k{special\_exit}, \k{promise\_force\_entry}, and
finally \k{promise\_force\_exit}.  The function-related callbacks are
used mainly for bookkeeping: the analysis is notified that a construct
has been entered by pushing the call onto a stack.  The calls
themselves store a trace object that holds the type information. As R
can perform single or multiple dispatch on function arguments
depending on their class, the relevant information is kept by the {\tt
  \_entry} variants.

\subsection{Checking signatures with contracts}

@@

\paragraph{User code}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
To represent end-user code in the corpus, we turned to Kaggle, an
online platform for data science and machine learning. The website
allows people to share data science competitions and data analysis
problems together with data for which users try to find the best
solution (something like a repository of hackathon or datathon
code). The solutions, called \emph{kernels}, are then posted on Kaggle
either as plain scripts or as notebooks. One of the most popular
competitions is about predicting passenger survival on
Titanic\footnote{\url{https://www.kaggle.com/c/titanic}} with
\KaggleDownloaded kernels in R (over 1/4 of all available R kernels)
which we used for our corpus.

@@

\paragraph{Type usage}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
During execution, 3,147 different types were observed.  Classes are the
most common types, accounting for roughly 31\% of types of arguments.
The most common classes are \c{matrices} (12\%), \c{data.frames}
(7.5\%), \c{formulas} (2\%), \c{factors} (2\%), and \c{tibbles} (2\%).
Roughly 25\% of classes are part of R's base libraries, the others are
user-defined.  Scalars and vectors are the next most common kind,
making up 41\% of remaining types. with scalars making up 28\% of
types and vectors 12\%.  Nulls and lists follow at 8\% and 7\%
respectively, and the vararg type makes up 6\% of arguments.  This all
totals up to over 90\% of types.  Table~\ref{ctb} reports on the 10
most frequent types occurring in the corpus. The first row of the
table reads: the \dbl type occurs in 12,298 (11.24\%) argument types,
and accounts for over 12 million (20.3\%) of the types observed by
\typetracer's dynamic analysis.

\begin{table}[!h]
% BEGIN Autogenerated
\small\begin{tabular}{lrrrr}\toprule
Type & Args & \% of Args & Observations & \% of Obs.\\\midrule
\dbl               & 12,298 &     11.24 &  12,152,787 & 20.3 \\
\lgl               & 9,366 &      8.7 & 6,650,294 & 11.1 \\
\nul              &  8,799 &      8.0 & 2,187,611 & 3.7 \\
\chr               &  8,727 &      8.0 & 2,564,207 & 4.3 \\
\VEC\dbl           &  7,190 &      6.6 &  4,934,773 & 8.2 \\
\VARGS             &  6,611 &      6.0 &  6,075,874 & 10.1 \\
\any               &  6,120 &      5.6 & 339,299 & 0.6 \\
\VEC\chr           &  4,325 &      4.0 &  1,060,466 & 1.8 \\
\CLASS{\texttt{matrix}}     & 4,152 & 3.8 & 2,805,718 & 4.7 \\
\CLASS{\texttt{data.frame}} & 2,608 & 2.4 &  352,655 & 0.6 \\
\bottomrule
\end{tabular}
\caption{Top types of arguments in R}\label{ctb}
\end{table}

@@

We ran \typetracer on the test, example, and vignette code of the
aforementioned corpus of \PACKAGES packages and successfully inferred
types for \NUMFUNCTIONS functions.  Table~\ref{tbl:typesig}
illustrates the process with ten representative signatures.  Many of
the features of our type language are represented here, and some
signatures are telling of the function's behaviour.  For example,
consider \code{decrypt_envelope}: the first three parameters of the
function are byte arrays, and the fourth argument is an RSA key, used
to decrypt some of the inputs, and the output of the function is
another byte array.  As another example, consider \code{Traverse}:
according to the function documentation, it takes the root of a tree
and traverses it in an order specified by the second argument.  We see
that reflected in the type, where the first argument has type
\CLASS{\texttt{Node}, \texttt{R6}} and the second argument had type
\VEC\chr, representing the traversal order.

\begin{table}[H]
  \centering
  \resizebox{.98\linewidth}{!}{
  \begin{tabular}{llrrr}
    \toprule
    Function & Type Signature \\
    \midrule
    \code{dplyr::group_indices} & \FUN{\CLASS{\texttt{data.frame}}, \VARGS}{\VEC\int}\\
    \code{moments::all.cumulants} & \FUN{\CLASS{\texttt{matrix}} \; | \; \VEC\dbl}{\CLASS{\texttt{matrix}}\; | \; \VEC{\dbl}}\\
    \code{diptest::dip}& \FUN{\VEC\dbl, \chr \; | \; \lgl, \lgl,
                         \dbl}{\CLASS{\texttt{dip}} \; | \; \dbl} \\
    \code{stabledist::cospi2}& \FUN{\VEC\dbl}{\VEC\dbl} \\
    \code{matrixcalc::matrix.power}& \FUN{\CLASS{\texttt{matrix}},
                                     \dbl}{\CLASS{\texttt{matrix}}} \\
    \code{data.tree::Traverse}& \FUN{\CLASS{\texttt{Node}, \texttt{R6}}, ~\VEC\chr,
                                \any, \any}{\LIST{\any}} \\
    \code{openssl::decrypt\_envelope}& \FUN{~\VEC\raw, ~\VEC\raw, ~\VEC\raw,
                                       \CLASS{\texttt{key}, \texttt{rsa}}, \any}{\VEC\raw} \\
    \code{dbplyr::set\_win\_current\_group}& \FUN{? ~\VEC\chr}{\; ? \; \VEC\chr} \\
    \code{openssl::sha256}& \FUN{~\VEC\raw, ? ~\VEC\raw}{\VEC\raw} \\
    \code{forecast::initparam}& \FUN{? \dbl, \any, \any, \any, \chr, \chr, \lgl,
                                ~\VEC\dbl, ~\VEC\dbl, \any}{\VEC\dbl} \\
    \bottomrule
  \end{tabular}}
\caption{Select Type Signatures}\label{tbl:typesig}
\end{table}

@@

\begin{table}
% BEGIN Autogenerated
\scriptsize
\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & Occurrences & \% Total & Cumul. \%\\
\midrule
\VEC\dbl & \CLASS{\texttt{matrix}} & 705,036 & 36.36 & 36.36 \\ 
\VEC\dbl & \dbl & 189,800 & 9.79 & 46.15 \\ 
\chr & \CLASS{\texttt{bignum}} | \VEC\raw & 100,100 & 5.16 & 51.31 \\
\CLASS{{\tt simpleError, error, condition}} & \makecell[tl]{\CLASS{{\tt data.frame}} | \CLASS{{\tt matrix}} | \\ \CLASS{{\tt randomForest}} | \VEC\dbl} & 78,197 & 4.03 & 55.34 \\ 
  \CLASS{{\tt data.frame}} & \CLASS{{\tt data.frame, tbl, tbl\_df}} & 58,809 & 3.03 & 58.37 \\ 
  \addlinespace
  \CLASS{{\tt matrix}} & \CLASS{{\tt timeSeries}} & 53,482 & 2.76 & 61.13 \\ 
 \VEC\dbl & \VEC\int & 33,350 & 1.72 & 62.85 \\ 
 \dbl & \CLASS{{\tt data.frame}} & 32261 & 1.66 & 64.52 \\ 
  \VEC\dbl & \CLASS{{\tt data.frame}} & 31361 & 1.62 & 66.13 \\ 
  \CLASS{{\tt xml\_node}} & \makecell[tl]{\CLASS{{\tt xml\_missing}} $|$ \CLASS{{\tt xml\_nodeset}} $|$ \\ \CLASS{{\tt xml\_document, xml\_node}}} & 30,330 & 1.56 & 67.70 \\ 
\bottomrule
\end{tabular}
% END Autogenerated
\caption{Top contract failures}
\label{tbl:fail-breakdown-0}
\end{table}

@@

 \begin{table}
% BEGIN Autogenerated
\scriptsize
\begin{tabular}{llrrr}
\toprule
              &                &      \multicolumn{2}{c}{\# Args Types}        &  \\
 Passed & Arg Type & Failed & Total & \% Failure\\
\midrule
\VEC\dbl & \CLASS{\texttt{matrix}} & 18 & 1522 & 1.18 \\ 
\VEC\dbl & \dbl  & 66 & 5865 & 1.13 \\ 
\chr & \CLASS{\texttt{bignum}} | \VEC\raw & 1 & 3 & 33.33 \\ 
\CLASS{{\tt simpleError, error, condition}} & \makecell[tl]{\CLASS{{\tt data.frame}} | \CLASS{{\tt matrix}} | \\ \CLASS{{\tt randomForest}} | \VEC\dbl}  & 2 & 2 & 100 \\ 
  \CLASS{{\tt data.frame}} & \CLASS{{\tt data.frame, tbl, tbl\_df}} & 23 & 196 & 11.73 \\ 
  \addlinespace
  \CLASS{{\tt matrix}} & \CLASS{{\tt timeSeries}} & 21 & 39 & 53.85 \\ 
 \VEC\dbl & \VEC\int & 31 & 624 & 4.97 \\ 
 \dbl & \CLASS{{\tt data.frame}} & 2 & 1025 & 0.20 \\ 
  \VEC\dbl & \CLASS{{\tt data.frame}}  & 6 & 1025 & 0.59 \\ 
  \CLASS{{\tt xml\_node}} & \makecell[tl]{\CLASS{{\tt xml\_missing}} $|$ \CLASS{{\tt xml\_nodeset}} $|$ \\ \CLASS{{\tt xml\_document, xml\_node}}}   & 1 & 1 & 100 \\ 
\bottomrule
\end{tabular}

% END Autogenerated
 \caption{Results from Table 4, broken down by occurrences of the expected type as a parameter type}
 \label{tbl:fail-breakdown-1}
 \end{table}

@@

Had we failed to capture some key usage pattern of R with our type
annotation framework, we would likely see it here. For example, consider Table~\ref{tbl:fail-breakdown-3}, which was
obtained identically to Table~\ref{tbl:fail-breakdown-2} except
selecting arguments in the 80th percentile instead.  The most frequent
argument type violation pattern in that of \CLASS{{\tt matrix}},
\VEC\dbl, and \CLASS{\texttt{data.frame}} values passed to arguments
expecting \CLASS{{\tt timeSeries}}.  This occurs in 35.90\% of such
arguments, and represents cases where tests did not adequately cover
all valid function inputs.  Separate from the issue of testing, we can
capture this behaviour with user-defined subtyping or coercion, as the
data types which were passed are readily convertible to \CLASS{{\tt
    timeSeries}}.

 \begin{table}
% BEGIN Autogenerated\
\scriptsize
\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
\midrule
\VEC\dbl & \VEC\int & 31 & 624 & 4.97 \\
\dbl & \int & 21 & 519 & 4.05\\
\VEC\chr & \chr | \nul & 10 & 256 & 3.91 \\
\NAVEC\lgl & \VEC\lgl & 8  & 219 & 3.65 \\
\NASCA\lgl & \VEC\lgl & 5 & 219 & 2.28 \\
\bottomrule
\end{tabular}
% END Autogenerated
 \caption{Highest failure rate among popular argument types, for
   argument signatures whose frequency is in the 90th percentile.}
 \label{tbl:fail-breakdown-2}
 \end{table}

  \begin{table}
% BEGIN Autogenerated
\scriptsize
\begin{tabular}{ll@{}rrr@{}}
\toprule
Passed & Arg Type & \#Args Fail & \#Args w Type & \% Failure\\
\midrule
\CLASS{{\tt matrix}} & \CLASS{{\tt timeSeries}} & 21 & 39 & 35.90 \\
\VEC\dbl & \CLASS{{\tt timeSeries}} & 21 & 39 & 35.90 \\
\CLASS{{\tt data.frame}} & \CLASS{{\tt timeSeries}} & 14 & 39 & 35.90 \\
\CLASS{{\tt dtplyr\_step\_first, dtplyr\_step}} & \makecell[tl]{\CLASS{{\tt data.frame}} | \\ \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}} | \\ \CLASS{{\tt data.frame, tbl, tbl\_df}}} & 10 & 45 & 22.22 \\
\chr & \raw & 6 & 31 & 19.35 \\
% \CLASS{\texttt{data.frame}} & \CLASS{\texttt{data.frame}, \texttt{tbl}, \texttt{tbl\_df}} & 63 & 284 & 22.18\\
% \CLASS{\texttt{matrix}} & \CLASS{\texttt{array}} & 13 & 70 & 18.57\\
% \NAVEC\lgl & \lgl | \nul & 9 & 49 & 18.37\\
% \VEC\chr & \LIST{\VEC\chr} & 20 & 109 & 18.35\\
% \VEC\clx & \clx & 7 & 42 & 16.67\\
\bottomrule
\end{tabular}
% END Autogenerated
 \caption{Highest failure rate among popular argument types, for argument signatures whose frequency is in the 80th percentile.}
 % *: Grouped Tibble is \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}} , and Tibble is \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}.}
 \label{tbl:fail-breakdown-3}
 \end{table}

>>>>

\begin{document}
\title{Designing Types for R, Empirically}
\author{Alexi Turcotte}\affiliation{\institution{Northeastern University, USA}}
\author{Aviral Goel}\affiliation{\institution{Northeastern University, USA}}
\author{Filip Křikava}\affiliation{\institution{Czech Technical University, Czechia}}
\author{Jan Vitek}\affiliation{\institution{Northeastern University, USA / Czech Technical University, Czechia}}

@@

\begin{abstract}
The R programming language is widely used in a variety of domains.  It
was designed to favor an interactive style of programming with minimal
syntactic and conceptual overhead. This design is well suited to data
analysis, but a bad fit for tools such as compilers or program
analyzers. In particular, R has no type annotations, and all
operations are dynamically checked at runtime. The starting point for
our work are the two questions: \emph{what expressive power is needed
to accurately type R code?} and \emph{which type system is the R
community willing to adopt?} Both questions are difficult to answer
without actually experimenting with a type system. The goal of this
paper is to provide data that can feed into that design process. To
this end, we perform a large corpus analysis to gain insights in the
degree of polymorphism exhibited by idiomatic R code and explore
potential benefits that the R community could accrue from a simple
type system.  As a starting point, we infer type signatures for
\NUMFUNCTIONS functions from \NUMPACKAGES packages among the most
widely used open source R libraries. We then conduct an evaluation on
\NUMPKGSEVAL clients of these packages, as well as on end-user code
from the Kaggle data science competition website.
\end{abstract} \maketitle

@@

R and its ancestor S were designed, implemented, and maintained by
statisticians. Originally they aimed to be glue languages for 
statistical routines written in Fortran. Over three decades they
became widely used across many fields for exploratory data analysis.
Modern R is fascinating as an object of study. It is a vectorized,
dynamically typed, lazy functional language with limited side-effects,
extensive reflective facilities and retrofitted object-oriented programming
support.

@@

The goal of this paper is to gather data that can be used as input to
the process of designing a type system for R.  To focus our work, we
chose to limit the scope of investigation on the design of a language
for function type signatures.  Thus, this paper will use the collected
data to document the signatures of user-defined functions.  For this,
we design a simple type language that matches the R data types.
We then extract call and return types from execution traces of a
corpus of widely used libraries, and finally synthesize type
signatures.  This allows us to see how far one can get with a simple
type language and identify limitations of our design. We validate the
robustness of the extracted type signatures by implementing a contract
checker that weaves types around their respective functions, and use
clients of the target packages for validation, as well as end-user code from data science competitions.

To sum up, we make the
following contributions:
\begin{itemize}
\item We implemented scalable and robust tooling to automatically
  extract type signatures and instrument R functions with checks
  based on their declared types.
\item We carried out a corpus analysis of \PACKAGES widely used and
  maintained packages to synthesize function type signatures and
  validated the robustness of the signatures against \NUMFILESRUN
  programs that use those functions.
\item We report on the appropriateness and usefulness of a simple type
  language for R.
\item Our data and code are open source and publicly
  available.\footnote{ \url{github.com/PRL-PRG/propagatr},
  \url{github.com/PRL-PRG/contractr}, data: \url{doi.org/10.5281/zenodo.4091818}, artifact: \url{doi.org/10.5281/zenodo.4037278}}
\end{itemize}

@@

\subsection{Synthesizing Signatures}

@@

\subsection{Types from Traces}

@@

\paragraph{Missing Arguments}
Parameters which receive no values when the function is called are
termed missing (not to be confused with \NA). This occurs when a
function is called with too few arguments and no default values are
specified for those missing arguments.  We record a \k{missing} type
for such arguments.  There are two obvious ways to deal with missing
arguments: type them as \any or type them as some unit type.  We
conservatively type them as \any.

\paragraph{Non-local Returns}
When a function exits with a longjump, there is no return value to speak
of. To ensure call traces are valid when a longjump occurs, we intercept the
unwinding process and record a special \k{jumped} return type for function
returns that are skipped. As we cannot be sure of the intended return value, 
these \k{jumped} values become \any types.

@@

\paragraph{Implementation Details.}
We primarily rely on eight callbacks: \k{closure\_entry},
\k{closure\_exit}, \k{builtin\_entry}, \k{builtin\_exit},
\k{special\_entry}, \k{special\_exit}, \k{promise\_force\_entry}, and
finally \k{promise\_force\_exit}.  The function-related callbacks are
used mainly for bookkeeping: the analysis is notified that a construct
has been entered by pushing the call onto a stack.  The calls
themselves store a trace object that holds the type information. As R
can perform single or multiple dispatch on function arguments
depending on their class, the relevant information is kept by the {\tt
  \_entry} variants.

\subsection{Checking Signatures with Contracts}

@@

\paragraph{User Code}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
To represent end-user code in the corpus, we turned to Kaggle, an
online platform for data science and machine learning. The website
allows people to share data science competitions and data analysis
problems together with data for which users try to find the best
solution (something like a repository of hackathon or datathon
code). The solutions, called \emph{kernels}, are then posted on Kaggle
either as plain scripts or as notebooks. One of the most popular
competitions is about predicting passenger survival on
Titanic\footnote{\url{https://www.kaggle.com/c/titanic}} with
\KaggleDownloaded kernels in R (over 1/4 of all available R kernels)
which we used for our corpus.

@@

\paragraph{Type Usage}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
During execution, 3,147 different types were observed.  Classes are the
most common types, accounting for roughly 31\% of types of arguments.
The most common classes are \c{matrices} (12\%), \c{data.frames}
(7.5\%), \c{formulas} (2\%), \c{factors} (2\%), and \c{tibbles} (2\%).
Roughly 25\% of classes are part of R's base libraries, the others are
user-defined.  Scalars and vectors are the next most common kind,
making up 41\% of remaining types. with scalars making up 28\% of
types and vectors 12\%.  Nulls and lists follow at 8\% and 7\%
respectively, and the vararg type makes up 6\% of arguments.  This all
totals up to over 90\% of types.  Table~\ref{ctb} reports on the 10
most frequent types occurring in the corpus. The first row of the
table reads: the \dbl type occurs in 12,298 (11.24\%) argument types,
and accounts for over 12 million (20.3\%) of the types observed by
\typetracer's dynamic analysis.

\begin{table}[!h]
\caption{Top types of arguments in R}\label{ctb}
% BEGIN Autogenerated
\small\begin{tabular}{lrrrr}\toprule
Type & Args & \% of Args & Observations & \% of Obs.\\\midrule
\dbl               & 12,298 &     11.24 &  12,152,787 & 20.3 \\
\lgl               & 9,366 &      8.7 & 6,650,294 & 11.1 \\
\nul              &  8,799 &      8.0 & 2,187,611 & 3.7 \\
\chr               &  8,727 &      8.0 & 2,564,207 & 4.3 \\
\VEC\dbl           &  7,190 &      6.6 &  4,934,773 & 8.2 \\
\VARGS             &  6,611 &      6.0 &  6,075,874 & 10.1 \\
\any               &  6,120 &      5.6 & 339,299 & 0.6 \\
\VEC\chr           &  4,325 &      4.0 &  1,060,466 & 1.8 \\
\CLASS{\texttt{matrix}}     & 4,152 & 3.8 & 2,805,718 & 4.7 \\
\CLASS{\texttt{data.frame}} & 2,608 & 2.4 &  352,655 & 0.6 \\
\bottomrule
\end{tabular}
\end{table}

@@

We ran \typetracer on the test, example, and vignette code of the
aforementioned corpus of \PACKAGES packages and successfully inferred
types for \NUMFUNCTIONS functions.  Table~\ref{tbl:typesig}
illustrates the process with ten representative signatures.  Many of
the features of our type language are represented here, and some
signatures are telling of the function's behavior.  For example,
consider \code{decrypt_envelope}: the first three parameters of the
function are byte arrays, and the fourth argument is an RSA key, used
to decrypt some of the inputs, and the output of the function is
another byte array.  As another example, consider \code{Traverse}:
according to the function documentation, it takes the root of a tree
and traverses it in an order specified by the second argument.  We see
that reflected in the type, where the first argument has type
\CLASS{\texttt{Node}, \texttt{R6}} and the second argument had type
\VEC\chr, representing the traversal order.

\begin{table}[H]
  \centering
  \caption{Select Type Signatures}\label{tbl:typesig}
  \resizebox{.98\linewidth}{!}{
  \begin{tabular}{llrrr}
    \toprule
    Function & Type Signature \\
    \midrule
    \code{dplyr::group_indices} & \FUN{\CLASS{\texttt{data.frame}}, \VARGS}{\VEC\int}\\
    \code{moments::all.cumulants} & \FUN{\CLASS{\texttt{matrix}} \; | \; \VEC\dbl}{\CLASS{\texttt{matrix}}\; | \; \VEC{\dbl}}\\
    \code{diptest::dip}& \FUN{\VEC\dbl, \chr \; | \; \lgl, \lgl,
                         \dbl}{\CLASS{\texttt{dip}} \; | \; \dbl} \\
    \code{stabledist::cospi2}& \FUN{\VEC\dbl}{\VEC\dbl} \\
    \code{matrixcalc::matrix.power}& \FUN{\CLASS{\texttt{matrix}},
                                     \dbl}{\CLASS{\texttt{matrix}}} \\
    \code{data.tree::Traverse}& \FUN{\CLASS{\texttt{Node}, \texttt{R6}}, ~\VEC\chr,
                                \any, \any}{\LIST{\any}} \\
    \code{openssl::decrypt\_envelope}& \FUN{~\VEC\raw, ~\VEC\raw, ~\VEC\raw,
                                       \CLASS{\texttt{key}, \texttt{rsa}}, \any}{\VEC\raw} \\
    \code{dbplyr::set\_win\_current\_group}& \FUN{? ~\VEC\chr}{\; ? \; \VEC\chr} \\
    \code{openssl::sha256}& \FUN{~\VEC\raw, ? ~\VEC\raw}{\VEC\raw} \\
    \code{forecast::initparam}& \FUN{? \dbl, \any, \any, \any, \chr, \chr, \lgl,
                                ~\VEC\dbl, ~\VEC\dbl, \any}{\VEC\dbl} \\
    \bottomrule
  \end{tabular}}
\end{table}

@@

\begin{table}
% BEGIN Autogenerated
\scriptsize
\caption{Top contract failures}
\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & Occurrences & \% Total & Cumul. \%\\
\midrule
\VEC\dbl & \CLASS{\texttt{matrix}} & 705,036 & 36.36 & 36.36 \\ 
\VEC\dbl & \dbl & 189,800 & 9.79 & 46.15 \\ 
\chr & \CLASS{\texttt{bignum}} | \VEC\raw & 100,100 & 5.16 & 51.31 \\
\CLASS{{\tt simpleError, error, condition}} & \makecell[tl]{\CLASS{{\tt data.frame}} | \CLASS{{\tt matrix}} | \\ \CLASS{{\tt randomForest}} | \VEC\dbl} & 78,197 & 4.03 & 55.34 \\ 
  \CLASS{{\tt data.frame}} & \CLASS{{\tt data.frame, tbl, tbl\_df}} & 58,809 & 3.03 & 58.37 \\ 
  \addlinespace
  \CLASS{{\tt matrix}} & \CLASS{{\tt timeSeries}} & 53,482 & 2.76 & 61.13 \\ 
 \VEC\dbl & \VEC\int & 33,350 & 1.72 & 62.85 \\ 
 \dbl & \CLASS{{\tt data.frame}} & 32261 & 1.66 & 64.52 \\ 
  \VEC\dbl & \CLASS{{\tt data.frame}} & 31361 & 1.62 & 66.13 \\ 
  \CLASS{{\tt xml\_node}} & \makecell[tl]{\CLASS{{\tt xml\_missing}} $|$ \CLASS{{\tt xml\_nodeset}} $|$ \\ \CLASS{{\tt xml\_document, xml\_node}}} & 30,330 & 1.56 & 67.70 \\ 
\bottomrule
\end{tabular}
% END Autogenerated

\label{tbl:fail-breakdown-0}
\end{table}

@@

 \begin{table}
% BEGIN Autogenerated
\scriptsize
 \caption{Results from Table 4, broken down by occurrences of the expected type as a parameter type}
\begin{tabular}{llrrr}
\toprule
              &                &      \multicolumn{2}{c}{\# Arg Types}        &  \\
 Passed & Arg Type & Failed & Total & \% Failure\\
\midrule
\VEC\dbl & \CLASS{\texttt{matrix}} & 18 & 1522 & 1.18 \\ 
\VEC\dbl & \dbl  & 66 & 5865 & 1.13 \\ 
\chr & \CLASS{\texttt{bignum}} | \VEC\raw & 1 & 3 & 33.33 \\ 
\CLASS{{\tt simpleError, error, condition}} & \makecell[tl]{\CLASS{{\tt data.frame}} | \CLASS{{\tt matrix}} | \\ \CLASS{{\tt randomForest}} | \VEC\dbl}  & 2 & 2 & 100 \\ 
  \CLASS{{\tt data.frame}} & \CLASS{{\tt data.frame, tbl, tbl\_df}} & 23 & 196 & 11.73 \\ 
  \addlinespace
  \CLASS{{\tt matrix}} & \CLASS{{\tt timeSeries}} & 21 & 39 & 53.85 \\ 
 \VEC\dbl & \VEC\int & 31 & 624 & 4.97 \\ 
 \dbl & \CLASS{{\tt data.frame}} & 2 & 1025 & 0.20 \\ 
  \VEC\dbl & \CLASS{{\tt data.frame}}  & 6 & 1025 & 0.59 \\ 
  \CLASS{{\tt xml\_node}} & \makecell[tl]{\CLASS{{\tt xml\_missing}} $|$ \CLASS{{\tt xml\_nodeset}} $|$ \\ \CLASS{{\tt xml\_document, xml\_node}}}   & 1 & 1 & 100 \\ 
\bottomrule
\end{tabular}

% END Autogenerated

 \label{tbl:fail-breakdown-1}
 \end{table}

@@

Had we failed to capture some key usage pattern of R with our type
annotation framework, we would likely see it here. For example, consider Table~\ref{tbl:fail-breakdown-3}, which was
obtained identically to Table~\ref{tbl:fail-breakdown-2} except
selecting arguments in the 80th percentile instead.  The most frequent
argument type violation pattern in that of \CLASS{{\tt matrix}},
\VEC\dbl, and \CLASS{\texttt{data.frame}} values passed to arguments
expecting \CLASS{{\tt timeSeries}}.  This occurs in 35.90\% of such
arguments, and represents cases where tests did not adequately cover
all valid function inputs.  Separate from the issue of testing, we can
capture this behavior with user-defined subtyping or coercion, as the
data types which were passed are readily convertible to \CLASS{{\tt
    timeSeries}}.

 \begin{table}
% BEGIN Autogenerated\
\scriptsize
 \caption{Highest failure rate among popular argument types, for
   argument signatures whose frequency is in the 90th percentile.}
\begin{tabular}{llrrr}
\toprule
Passed & Arg Type & \# Args Failed & \# Args with Type & \% Failure\\
\midrule
\VEC\dbl & \VEC\int & 31 & 624 & 4.97 \\
\dbl & \int & 21 & 519 & 4.05\\
\VEC\chr & \chr | \nul & 10 & 256 & 3.91 \\
\NAVEC\lgl & \VEC\lgl & 8  & 219 & 3.65 \\
\NASCA\lgl & \VEC\lgl & 5 & 219 & 2.28 \\
\bottomrule
\end{tabular}
% END Autogenerated

 \label{tbl:fail-breakdown-2}
 \end{table}

  \begin{table}
% BEGIN Autogenerated
\scriptsize
 \caption{Highest failure rate among popular argument types, for argument signatures whose frequency is in the 80th percentile.}
\begin{tabular}{ll@{}rrr@{}}
\toprule
              &                &      \multicolumn{2}{c}{\# Arg Types}        &  \\
 Passed & Arg Type & Failed & Total & \% Failure\\
\midrule
\CLASS{{\tt matrix}} & \CLASS{{\tt timeSeries}} & 21 & 39 & 35.90 \\
\VEC\dbl & \CLASS{{\tt timeSeries}} & 21 & 39 & 35.90 \\
\CLASS{{\tt data.frame}} & \CLASS{{\tt timeSeries}} & 14 & 39 & 35.90 \\
\CLASS{{\tt dtplyr\_step\_first, dtplyr\_step}} & \makecell[tl]{\CLASS{{\tt data.frame}} | \\ \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}} | \\ \CLASS{{\tt data.frame, tbl, tbl\_df}}} & 10 & 45 & 22.22 \\
\chr & \raw & 6 & 31 & 19.35 \\
% \CLASS{\texttt{data.frame}} & \CLASS{\texttt{data.frame}, \texttt{tbl}, \texttt{tbl\_df}} & 63 & 284 & 22.18\\
% \CLASS{\texttt{matrix}} & \CLASS{\texttt{array}} & 13 & 70 & 18.57\\
% \NAVEC\lgl & \lgl | \nul & 9 & 49 & 18.37\\
% \VEC\chr & \LIST{\VEC\chr} & 20 & 109 & 18.35\\
% \VEC\clx & \clx & 7 & 42 & 16.67\\
\bottomrule
\end{tabular}
% END Autogenerated
 % *: Grouped Tibble is \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}} , and Tibble is \CLASS{{\tt data.frame, grouped\_df, tbl, tbl\_df}}.}
 \label{tbl:fail-breakdown-3}
 \end{table}
